Computational Methods for Principle Component Analysis Applied to Genotype Data

### 2018-06-20 doc/paper/paper.tex
Finish writing results

\usepackage{epsfig}
\usepackage[margin=1in]{geometry}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[capitalise]{cleveref}
\usepackage{verbatim}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}

\newcommand{\bO}{\mathcal{O}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\ostu}{\operatorname{stu}}
\newcommand{\oref}{\operatorname{ref}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}

@@

\begin{abstract}
    Population stratification (PS) is an important confounder
    for the association between genotypes and phenotypes.
    To adjust for PS,
    we need to accurately estimate the ancestral structure
    in the study samples.
    This can be done effectively
    by using principle component (PC) analysis
    to predict the PC scores of the study genotypes
    from a set of reference genotypes.
    There are two major existing methods for predicting PC scores:
    simple projection (SP) and augmentation-decomposition-transformation (ADP).
    However, when the reference size is large,
    the former is computationally efficient but inaccurate
    (PC scores shrinking toward zero),
    while the latter is accurate but computationally costly ($\bO[n^3]$).
    To solve these problems,
    we created the genoPCA software
    that implemented 
    the adjusted projection (AP) method to remove the bias in SP, and
    the online ADP (OADP) method to speed up ADP.
    We compared the accuracy and computational efficiency of these four methods
    by conducting a simulation study
    with the GGS software.
    We also used them to study the UKbiobank data by using the 1000Genomes data as the reference.
    We found that AP achieved the accuracy of ADP without adding much runtime,
    while OADP reduced the runtime of ADP by $x$-fold without losing accuracy.
    Furturmore, when studying a large number of samples
    from a narrower geographic area
    where the signal-to-noise ratio is low,
    AP and OADP performed reasonably well
    while SP frequently misclassified the populations
    and the runtime of ADP became impractical.

@@

We used the Grid Genotype Simulator (GGS)
by \citet{ mathieson2012differential}
to simulate population migration.
The grid's dimension is 2 by 2
with the same number of individuals
in each of the 4 populations.
Moreover, we let the number of reference samples
range from 1000 to 3000
and fix the number of study samples to 200.
Each genotype sample has 100 genealogies
with 1000 loci in each genealogy.
In addition, the migration rate was set to 100.
After the individuals were simulated,
we applied SP, ADP, AP, and OADP to the data
to predict the PC scores for the study samples.
Then we used the k-nearest neighbors method
to predict which ethnic group a study sample comes from.

@@

We also applied the four methods
to study the UK Biobank data \citep{sudlow2015uk},
which consists of 490,000 middle- or elderly-aged individuals
in the United Kingdom.
Our reference samples
are from the 1000 Genomes Project \citep{10002015global},
which collected 2500 samples from Europe, Africa, East Asia, South Asia, and America.
(Individuals whose children are also in the data were removed.)
The reference and the study samples
have 145,000 SNPs in common.
After we predicted the PC scores and ethnic groups
of the study samples,
we selected the study samples predicted to by Europeans
and applied PCA to them
by using the European reference samples only.
Since ADP is very slow compared to the other three methods,
we randomly selected 500 study samples
and used them to compare the four methods.
After that,
we applied SP, AP, and OADP to all the study samples.

@@

\subsection{Simulation study}

% Runtimes vs ref size
% Distance to ADP vs ref size
% Distance to geo center vs ref size
% PC1 vs PC2, 5 x 3 grid (uncolored study samples)
% The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.

@@

For computational cost,
we can see that the runtime for SP increases faster than linearly
as the reference sample size increases.
This is consistent
with the theoretical computational complexity of SP,
which expects the runtime to increase at the rate of $\bO[n^3]$.
In comparison, the runtimes of SP, AP, and OADP
increase only slightly.
The theoretical computational complexity of SP and AP is $\bO[1]$
and that of OADP is $\bO[n]$,
when the data dimension and the study sample size are fixed.

@@

When we used the global samples in the UK Biobank data,
we found that the PC scores predicted by the four methods' are almost identical.
The separation of the ethnic groups was clear,
and the shrinkage bias of SP was not obvious,
which is consistent with the theoretical understanding
that the shrinkage bias decreases
as the signal-to-noise ratio increases.

However, when we focused on the European (EUR) samples only,
the results given by SP were clearly unreliable.
The samples from British in England and Scotland (GBR)
and those from Utah residents with northern and western European ancestry	(CEU)
drifted very close to NULL.
This causes SP to incorrectly classified almost all of them as CEU,
since most GBR reference samples are far away from NULL.
A few of the study samples predicted by ADP to be GBR or CEU
were even classified as the Iberian population in Spain (IBS),
since the IBS reference samples are close to NULL.
Moreover, the shrinkage for the IBS
and for the Toscani in Italia (TSI) populations
caused significant errors within these two groups.
The IBS reference samples
are approximately in the middle
between the TSI reference samples and NULL.
Thus most of the the study samples predicted by ADP to be TSI
were classified by SP as IBS.

For OADP, its predicted PC1 and PC2 scores
for the European study samples
are almost identical
to those done by ADP.
OADP's results for PC3 and PC4
are also very close to ADP's results,
though we observed a very slight decrease in the magnitude of the PC scores.
This minor shrinkage did not affect
OADP's classification for the IBS and TSI samples,
since their PC1 and PC2 scores were accurately predicted by OADP.
Neither was any CEU/GBR sample classified as IBS/TSI. 
For the CEU and GBR samples,
their predicted PC scores are tightly mixed together
for all the 4 PCs,
which were even difficult for ADP to separate them.
Thus the slight difference in PC3 and PC4 scores
between OADP and ADP
caused the classification of many CEU and GBR samples
to be different on the two methods.

Furthermore, AP successfully adjusted for the shrinkage
in PC1 and PC2 for the European samples,
so its predicted PC scores are almost identical to ADP's.
However, it did not detect the shrinkage in PC 3 and PC4,
so its predicted PC scores are the same as SP's for these two PCs.
This still allowed AP to correctly separate
the CEU/GBR samples from the IBS/TSI samples.
However, its separation of the IBS and TSI samples
did not agree completely with that done by ADP,
most likely caused by the shrinkage on PC3.
For the classification between CEU and GBR,
the AP's results differ far from ADP's,
as AP classified many more CEU/GBR samples as CEU.

The patterns for the accuracy of SP, AP, and OADP
discussed above
were also observed
when we used all the 490,000 study samples in the UK Biobank dataset
and those predicted to be Europeans among them.
See \Cref{tbl:ukb} for details.

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    \multicolumn{6}{|c|}{Study Runtime} \\
    \hline
    SP         &  17  & 17   & 16   & 18   & 18   \\
    AP         &  18  & 16   & 18  & 16   & 17  \\
    OADP       &  145  & 146   & 147   & 148   & 149   \\
    ADP        &  379  & 809 & 1526 & 2556 & 4062 \\
    \hline
    \multicolumn{6}{|c|}{Distance from ADP's results} \\
    \hline
    SP         &  188  & 139   & 112   & 91   & 84   \\
    AP         &  31  & 27   & 12  & 11   & 14  \\
    OADP       &  21  & 16   & 12   & 9   & 10   \\
    ADP        &  0  & 0 & 0 & 0 & 0 \\
    \hline
    \multicolumn{6}{|c|}{Distance from reference centers} \\
    \hline
    SP         &  335 & 161  & 113 & 71  & 60  \\
    AP         &  31  & 19   & 17  & 25  & 15  \\
    OADP       &  12  & 17   & 11  & 20  & 12  \\
    ADP        &  20  & 14   & 14  & 23  & 14  \\
    \hline
  \end{tabular}
  \caption{
    The runtimes and errors of the four methods
    as the reference size increases.
    Distances are measured by the Euclidean distance.
    Reference centers are the centers of mass of the populations
    in the reference samples.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
    }
\end{table}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{|r|r|r|r|r|r|r|}
    \hline
    Method & \multicolumn{2}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
    \hline
    & Global & European & Global & European & Global & European \\
    \hline
    SP         &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
    AP         &  65   & 48  & 48  & 33  & 2486 & 706  \\
    OADP       &  363  & 349 & 10  & 32  & 2575 & 665  \\
    ADP        &  6739 & 518 & 0   & 0   & 2578 & 649  \\
    \hline
  \end{tabular}
  \caption{
    The study runtimes (in seconds) and erros for the four methods
    for 500 randomly selected study samples
    in the UK Biobank data.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference samples
    to the centers of mass of the study samples.
    For the global study,
    2492 samples from the 1000 Genomes data are used as the reference.
    For the European study,
    only the 498 European samples among them are used.
    The number of variants in common is 145,282.
    Only the top 4 PCs are calculated.
    }
\end{table}

>>>>

\usepackage{epsfig}
\usepackage[margin=1in]{geometry}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[capitalise]{cleveref}
\usepackage{verbatim}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{color}

\newcommand{\bO}{\mathcal{O}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\ostu}{\operatorname{stu}}
\newcommand{\oref}{\operatorname{ref}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
\newcommand{\todo}[1]{{\color{red} (#1)}}

@@

\begin{abstract}

 Population stratification (PS) is a major confounder in genome-wide association studies (GWAS) and can lead to false positive associations. To adjust for PS, principal component analysis (PCA)-based ancestry prediction has been widely used. This approach first performs PCA on the reference samples with detailed ancestry information and then uses the reference sample PCA results to predict the PC scores of the study samples. Simple Projection (SP) and Augmentation-Decomposition-Transformation (ADP), such as LASER and TRACE, are popular methods for predicting PC scores. However, they are either biased or computationally expensive. For example, the predicted PC scores from SP can be biased toward NULL. On the other hand, since ADP requires running PCA separately for each study sample on the augmented dataset, its computational cost is high.

To address these problems, we propose two alternative approaches, Adjusted Projection (AP) and Online ADP (OADP), implemented in the genoPCA software. Using random matrix theory results, AP asymptotically estimates and adjusts for the bias of SP. Moreover, OADP uses online singular value decomposition, which can greatly reduce the computation cost of ADP. We carried out extensive simulation studies to show that these alternative approaches are unbiased and that the computation times are 10-100 times faster than ADP when 2000 reference samples are used. We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference. AP and OADP required 16 and 92 CPU hours, respectively, while the projected computation time of ADP is 1,933 CPU hours. Furthermore, when we only used the European reference samples in the 1000 Genomes to infer sub-European ancestry, SP clearly showed bias, unlike the proposed approaches. By using AP and OADP, we can infer ancestry and adjust for PS in GWAS robustly and efficiently

@@

We used the Grid Genotype Simulator (GGS)
by \citet{mathieson2012differential}
to simulate population migration.
The grid's dimension is 2 by 2
with the same number of individuals
in each of the 4 populations.
Moreover, we let the number of reference samples
range from 1000 to 3000
and fix the number of study samples to 200.
Each genotype sample has 100 genealogies
with 1000 loci in each genealogy.
In addition, the migration rate was set to 100.
After the individuals were simulated,
we applied SP, ADP, AP, and OADP to the data
to predict the PC scores for the study samples.
We only calculated the top 4 PCs,
and for OADP and ADP,
we calculated the top 20 PC scores for the study samples
and projected them to the dim-4 reference PC score space
with a Procrustes transformation.
Finally, we used the k-nearest neighbors method
to predict which ethnic group a study sample comes from.

We used two measures for the accuracy of PC score prediction.
First, we measured the ADP error,
which is the Euclidean distance
from a method's predicted PC scores
to ADP's PC scores,
since ADP is believed to be the currently most reliable method.
Next, for the center error,
we calculated the centers of mass of each population's PC scores.
This was done for both the reference and the study samples.
Then the center error
was set to be the Euclidean distance
from the reference centers to the corresponding study centers.

For the comparison of computational cost,
we recorded each method's runtime for analyzing the study samples.
(The runtime for PCA on the reference samples
and for reading or writing files are not included.)
All the methods were run on a single CPU core.

@@

We also applied the four methods
to study the UK Biobank data \citep{sudlow2015uk},
which consists of 488,000 middle- or elderly-aged individuals
in the United Kingdom.
Our reference samples
are from the 1000 Genomes Project \citep{10002015global},
which collected 2500 samples from Europe, Africa, East Asia, South Asia, and America.
(Individuals whose children are also in the data were removed.)
The reference and the study samples
have 145,000 SNPs in common.
After we predicted the PC scores and ethnic groups
of the study samples,
we selected the study samples predicted to by Europeans
and applied PCA to them
by using the European reference samples only.
Since ADP is very slow compared to the other three methods,
we randomly selected 500 study samples
and used them to compare the four methods.
After that,
we applied SP, AP, and OADP to all the study samples.
The same measurements for accuracy and computational cost
from the simulation study
was applied here.

@@

\subsection{Simulation studies}

@@

The accuracy of the four methods were compared
by using their ADP errors and center errors.
Compared to SP,
the other three methods' errors are very small
and close to each other.
OADP had smaller errors than AP
for both the ADP errors and the center errors.
Interestingly, the center errors of ADP were slightly higher
than that of OADP.
As the reference size increased,
the errors of all the four methods decreased.
SP exhibited the fastest decrease in both errors,
and the four methods' difference in accuracy
became smaller and smaller.

For computational cost,
we observed that the runtime for ADP increases faster than linearly
as the reference sample size increases.
This is consistent
with the theoretical computational complexity of ADP,
which expects the runtime to increase at the rate of $\bO[n^3]$.
When the reference sample size reached 3000,
ADP's runtime was 27 times of OADP's
and 225 times of SP's and AP's.
In comparison, the runtimes of SP, AP, and OADP
increased only slightly,
which is consistent with their $\bO[1]$ and $\bO[n]$ computational complexity when the data dimension and study sample size are fixed.
SP and AP were the most computational efficient methods.
Their runtimes were approximately only one-eighth of OADP's.

@@

When we used the 500 randomly selected global samples
in the UK Biobank data,
we found that the PC scores predicted by the four methods' are almost identical.
The separation of the ethnic groups was clear,
and the shrinkage bias of SP was not obvious,
which is consistent with the theoretical understanding
that the shrinkage bias decreases
as the signal-to-noise ratio increases \todo{Find citation}.

However, when we focused on the European (EUR) samples
among these 500 UK Biobank samples,
(of which there are 475 many,)
the results given by SP were clearly inaccurate.
The samples from British in England and Scotland (GBR)
and those from Utah residents with northern and western European ancestry	(CEU)
drifted very close to NULL.
This caused SP to incorrectly classified almost all of them as CEU,
since most GBR reference samples are farther away from NULL.
A few of the study samples predicted by ADP to be GBR or CEU
were even classified as the Iberian population in Spain (IBS),
since the IBS reference samples are close to NULL.
Moreover, the shrinkage for the IBS
and for the Toscani in Italia (TSI) populations
caused significant errors within these two groups.
The IBS reference samples
are approximately in the middle
between the TSI reference samples and NULL.
Thus most of the the study samples predicted by ADP to be TSI
were classified by SP as IBS.
The same patterns for the accuracy
of SP, AP, and OADP
discussed above
were observed
when we used all the 488,000 study samples in the UK Biobank dataset
and those predicted to be Europeans among them.

For AP and OADP, their predicted PC1 and PC2 scores
for the European study samples
were almost identical
to those done by ADP.
The three methods' classification
agreed on most of the samples,
except for within the cluster of CEU and GBR.
Since many residents with northern and western European ancestry living in Utah
could originally migrated from Britain,
these two groups' ancestral history could be very similar
and thus difficult if not impossible to be separated by PCA.
Hence any slightest change
in the PC scores for these two groups' study samples
could easily change the classification from one group to the other.
This is a problem inherent to the population structure
and is not caused by the limitations of our methods.

Just as in the simulation study,
SP's errors were the greatest.
Its ADP error
was twice of SP's and 9 times of OADP's for the global samples,
and 14 times of SP's and OADP's for the European samples.
For the center errors,
SP's error was still much greater than the other three methods.
For the European samples,
ADP's center error was the lowest,
followed by OADP and AP.
Interestingly, this order of the three methods
was reversed for the global samples.

For the computational cost,
ADP's runtime was 1.5 times of OADP's and 10 times of AP's and SP's
when only the European samples
of the 500 random global study samples
and the 498 European reference samples of the 1000 Genomes data
were used.
This difference increased to 18 times, 103 times, and 108 times
when we studied the all the 500 random global samples
with all the 2492 1000 Genomes reference samples.
As we included all the 488,000 study samples in the UK Biobank data,
SP, AP, and OADP only took 13, 16, and 90 CPU hours,
while ADP was estimated (from its 500-sample runtime)
to take 1828 CPU hours,
which makes ADP very computational costly to use in practice.

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    \multicolumn{6}{|c|}{Study Runtime} \\
    \hline
    SP         &  17  & 17   & 16   & 18   & 18   \\
    AP         &  18  & 16   & 18  & 16   & 17  \\
    OADP       &  145  & 146   & 147   & 148   & 149   \\
    ADP        &  379  & 809 & 1526 & 2556 & 4062 \\
    \hline
    \multicolumn{6}{|c|}{Error (ADP)} \\
    \hline
    SP         &  188  & 139   & 112   & 91   & 84   \\
    AP         &  31  & 27   & 12  & 11   & 14  \\
    OADP       &  21  & 16   & 12   & 9   & 10   \\
    ADP        &  0  & 0 & 0 & 0 & 0 \\
    \hline
    \multicolumn{6}{|c|}{Error (centers)} \\
    \hline
    SP         &  335 & 161  & 113 & 71  & 60  \\
    AP         &  31  & 19   & 17  & 25  & 15  \\
    OADP       &  12  & 17   & 11  & 20  & 12  \\
    ADP        &  20  & 14   & 14  & 23  & 14  \\
    \hline
  \end{tabular}
  \caption{
    The runtimes and errors of the four methods
    as the reference size increases.
    Runtimes are measured in seconds.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference samples
    to the centers of mass of the study samples.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
    }
\end{table}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{|r|r|r|r|r|r|r|r|}
    \hline
    & \multicolumn{3}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
    \hline
    Samples & All (hrs) & Global (sec) & European (sec) & Global & European & Global & European \\
    \hline
    Sample size & 488,000 & 500 & 475 & 500 & 475 & 500 & 475 \\
    \hline
    SP   & 13    &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
    AP   & 16    &  65   & 48  & 48  & 33  & 2486 & 706  \\
    OADP & 90    &  363  & 349 & 10  & 32  & 2575 & 665  \\
    ADP  & 1828*  &  6739 & 518 & 0   & 0   & 2578 & 649  \\
    \hline
  \end{tabular}
  \caption{
    The study runtimes and erros for the four methods
    for 1) all the samples and 2) 500 randomly selected study samples
    in the UK Biobank data.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference populations
    to the corresponding centers of mass of the study populations
    predicted by the $k$-nearest neighbor method.
    For the global study,
    2492 samples from the 1000 Genomes data are used as the reference.
    For the European study,
    only the 498 European samples among them are used.
    The number of variants
    in common between the reference and study samples is 145,282.
    Only the top 4 PCs are calculated.
    *Estimated from the runtime of ADP for the 500 global samples.
    }
\end{table}
